---
title: "I. Parting Thoughts"
author: Jeffrey W. Hollister
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  rmarkdown::html_vignette:
    toc: true
    fig_width: 6
    fig_height: 6
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{I. Parting Thoughts}
  \usepackage[utf8]{inputenc}
---


```{r setup, echo=FALSE}
title="I. Parting Thoughts"
gsIntroR::navigation_array(title)
```

Having now presented these materials to multiple groups of eager USGS students, I wanted to add one last post to the site with some parting thoughts.  There are many ways to successfully complete the exercises so your code will very likely be different than what I have presented here.  That being said the answers should not be different (except randomForest).

##Parting Thoughts
It is impossible to become proficient at anything over the course of two days and R certainly is not an exception to this.  What I do hope this workshop and/or website has provided is a starting point.  The only way to get better with R is to keep using it.  The best way I know of to do this is to keep reading, taking workshops, and trying to solve problems with R.  A diversity of approaches is not only a good idea but with something as broad as R I think it required.  And, you will always pick up new tricks.  

In addition to an immersion approach, another REALLY good thing to do is to find a problem that you need to solve and force yourself to solve it with R.  I know most everyone has had many years experience with data analysis with other software.  When you have this background, it is certainly easy to fall back on what you know, but use this opportunity to not only get something done, but make sure you allow additional time to also learn R.  A warning though, I would not recommend going cold turkey and just diving into R.  Ramp up slowly over the next year or so and try to keep solving one more problem with R.  This allows you to keep working while also slowly building your profeciency.  In short, slow and steady.  

```{r echo=FALSE}
gsIntroR::navigation_array(title)
```
